Simulating physiological flexibility in the acute glucocorticoid response to stressors reveals limitations of current empirical approaches

### 2022-06-30 3_r_scripts/Revision.Rmd
Working on revision.

---
title: "Simulating physiological flexibility in the acute glucocorticoid response to stressors reveals limitations of current empirical approaches"
author: 
 - Conor C. Taff
header-includes:
  - \usepackage[font={footnotesize}, labelfont={bf}]{caption}
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}\linenumbers
fontsize: 12pt
output: 
  word_document:
  bookdown::pdf_document2:
      number_sections: FALSE
      toc: FALSE
      extra_dependencies: ["flafter"]
  bookdown::word_document2:
      number_sections: FALSE
      toc: FALSE
      extra_dependcies: ["flafter"]
bibliography: references.bib
csl: animal-behaviour.csl
---

@@

Wild animals often experience unpredictable challenges that demand rapid and flexible responses. The glucocorticoid mediated stress response is one of the major systems that allows vertebrates to rapidly adjust their physiology and behavior. Given its role in responding to challenges, evolutionary physiologists have focused on the consequences of between-individual and, more recently, within-individual variation in the acute glucocorticoid response. However, empirical studies of physiological flexibility are severely limited by the logistical challenges of measuring the same animal multiple times. Data simulation is a powerful approach when empirical data are limited, but has not been adopted to date in studies of physiological flexibility. In this paper, I develop a simulation that can generate realistic acute glucocorticoid response data with user specified characteristics. Simulated animals can be sampled continuously through an acute response and across as many separate responses as desired, while varying key parameters. Using the simulation, I develop several scenarios that address key questions in physiological flexibility. These scenarios demonstrate the conditions under which a single glucocorticoid trait can be accurately assessed with typical experimental designs, the consequences of covariation between different components of the acute stress response, and the way that context specific differences in variability of acute responses can influence the power to detect relationships between the strength of the acute stress response and fitness. I also describe how to use the simulation tools to aid in the design and evaluation of empirical studies. Recently there has been a great deal of interest in understanding the causes and consequences of physiological flexibility, but empirical data are often extremely limited and traditional sampling methods are poorly designed to directly address the questions of interest. This simulation represents a way forward by revealing critical aspects of physiological flexibility and by creating a tool for designing better empirical studies that integrate simulation and theory with data collection.

@@

Animals live in a dynamic environment in which they regularly encounter unpredictable challenges. Successfully navigating these challenges often requires the ability to rapidly adjust behavior and physiology to match current conditions. For vertebrates, the glucocorticoid mediated stress response plays a major role in coordinating these changes when stressors are encountered [@wingfield1998; @sapolsky2000] and similar rapid response systems mediate changes in other taxa [@taborsky2020]. Because of the central role that this response plays in coping with challenges, a great deal of research effort over the past 15 years has focused on understanding whether between-individual differences in the magnitude of this response predict coping ability and, ultimately, fitness [@breuner2008; @schoenle2020].

@@

Several recent papers have suggested that physiologists interested in endocrine flexibility should adopt a within-individual reaction norm approach [e.g., @hau2016; @taff2016]. This approach has been widely adopted in studies of behavioral flexibility where statistical methods and empirical progress have developed synergistically [e.g., @araya2015; @dingemanse2010; @westneat2015]. This field has also benefited from simulation studies to evaluate optimal study design [@van2012] and packages that can create artificial datasets with desired patterns of between, within, and residual variance to evaluate the consequences of different patterns of variation on the ability to detect effects [see SQuID package, @allegue2017]. While these approaches are powerful, they have proven difficult to apply directly to endocrine flexibility data for two reasons. First, simulation studies suggest that many patterns may only be detectable with a level of repeated sampling that is possible for many behaviors (especially when collected autonomously), but that is currently not possible for most studies of endocrine flexibility. Second, and more fundamentally, these papers often focus on somewhat discrete measures of behavior (e.g., aggression score or activity level), whereas for acute glucocorticoid responses, the functional shape of the response itself may be the important trait and it may not be possible to summarize variation in the shape of the response with a single measure.

The function valued trait (FVT) framework is an alternative approach that explicitly considers the functional shape of a biological response [@gomulkiewicz2018; @stinchcombe2012; @kingsolver2015]. While FVT approaches have been suggested for studies of endocrine flexibility [@taff2016], I am not aware of any papers that have applied this framework to empirical data on acute glucocorticoid responses, probably because sufficient data are not available. Conceptually, however, this approach is a better match to the acute glucocorticoid response, because the shape of a response curve is explicitly considered as the phenotypic trait of interest. In some cases, it may make sense to estimate particular parameters of the curve (e.g., maximum rate of increase and maximum value reached) and then treat those parameters as phenotypic values for downstream analysis, although statistical methods also exist to analyze the shape of the entire curve directly without the need to extract discrete parameters [@kingsolver2015]. This approach has been used to study a variety of phenotypes where values can be measured continuously or pooled across many individuals from the same group to accurately estimate the shape of a curve [see Table 1 in @stinchcombe2012]. Applying the technique to endocrine flexibility at the within-individual level faces the same empirical challenges described for within-individual reaction norms above. Note that FVT and within-individual reaction norms approaches are not necessarily incompatible, but they have largely developed separately.

@@

For individual based studies, the most common approach to this problem is to standardize measurements as much as possible by measuring animals at the same time of the day during the same context, and by taking blood samples at standard times (often <3 and 30 minutes after capture) to characterize baseline and stress-induced glucocorticoids. This standardization allows for comparison between individuals, but in some cases it may also completely obscure the ability to detect variation in certain characteristics of the acute response curve. For example, if the speed (rate of initial increase) and scope (maximum value) of the acute response vary independently, samples taken at only two time points cannot accurately capture variation in either parameter. Indeed, several discussions in recent years about methods such as the '3 minute rule' and the relative merits of 'area under the curve' versus time point measures of glucocorticoids are fundamentally related to a recognition of the importance of understanding variation in the functional shape of stress responses and whether different components of that shape covary within individuals [e.g., @cockrem2002; @small2017].

One of the characteristics of both the within-individual reaction norm and FVT literature is that empirical work has proceeded in very close coordination with simulation and statistical method development. In contrast, studies of endocrine flexibility often point to these methods, but don't address the ways that the particular logistical challenges of hormone measurement might necessitate different empirical approaches. I believe this is one reason that there are currently more conceptual papers arguing for a reaction norm approach to endocrine variation than there are empirical papers actually applying the approach. While many of the tools developed in these related fields are transferable, studies of physiological flexibility would benefit from a focus on analysis development and testing that explicitly incorporates the particular details and challenges of these questions. One way to accomplish these goals is to use simulations, but to my knowledge no studies of physiological flexibility have developed simulations of the acute stress response that address the issues discussed above.

@@

@@

I developed a set of functions in R version 4.0.2 [@rcore] to generate acute physiological response curves. This simulation makes no assumptions about the  mechanistic process that results in the shape of a glucocorticoid response. Rather, parameters are sampled to generate curves that are similar in shape and degree of variation to empirically observed responses (Figure \@ref(fig:concept-fig)). This simulation is designed to create data sets with realistic structure that can be used to better design and plan studies of physiological flexibility, to evaluate power of current study designs, and to evaluate the sensitivity of sampling regimes to any number of modifications to the shape of glucocorticoid response curves (e.g., changing covariation patterns between different features of the response). I explore a small number of scenarios in the next section, but I expect that many other scenarios can be addressed with these tools. For illustration purposes, I refer to simulated glucocorticoid responses, but the simulation applies equally well to any physiological mediator of a rapid response. The package can be installed in R using the following command.

@@

The simulation is constructed as two main functions with several minor functions for downstream analysis. Detailed descriptions of the arguments to each function are included with the package documentation. Briefly, function `cort_sim1` samples the parameters shown in Figure \@ref(fig:concept-fig) from an arbitrary number of animals. These parameters are sampled from a multivariate normal distribution with user specified mean, variance, and covariance for each parameter. I consider these values to be the 'true', unobserved, phenotype of the animal (setting aside the question of whether or not a 'true' physiological phenotype exists).

A second function, `cort_sim2`, starts with a population of animals generated from `cort_sim1` and samples observed acute glucocorticoid responses an arbitrary number of times for each animal. Two sources of variation in the observed relative to true parameter values can be specified. First, within-individual variation in expression is represented by specifying what amount of variation in the observation of each parameter is determined by the true value and what amount is determined by an additional randomly sampled response, based on the population parameters (this additional sampling maintains the user specified covariance structure of the population). After sampling the parameters, values are interpolated for each one minute time point and a localized regression is fit to create a smoothed curve that represents the observed glucocorticoid response. From this expressed response, individual data points are then collected at user specified times that would reflect an empirical study design (e.g., 2, 30, and 60 minutes). Additional noise can be added to these data points to represent measurement error (e.g., assay error).

The function also generates a simulated performance (e.g., fitness) measure, based on the underlying true values. Data reflecting the true phenotypic values, the repeated expression of acute responses, and the observed time points can then be used in downstream analyses with any standard statistical approaches or software. For example, a user could perform an analysis to ask whether a known relationship between fitness and a particular true parameter is recovered in a study that includes only measures taken at particular time points. An additional convenience function summarizes the output of a simulation run in a multi-panel plot (Figure \@ref(fig:initial-demo)).

```{r initial-demo, results = FALSE, echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10.2, fig.height = 7.5, fig.cap = "Example of simulation output with default settings. Panel A shows the downsampled data set for this run with samples collected at 1, 15, and 30 minutes in this case. Panel B shows the full observed response curve for each animal. Panel C shows the rank order of glucocorticoid level at each time point for each animal. In each panel, the vertical dashed lines represent the three time points that might have been measured in a typical empirical study. Note that individuals in the top panels do not match perfectly because measurement error is added to the downsampled dataset in panel A."}
  # Make plot with default settings from package 
     set.seed(100)   # makes it reproducible by using same random seed
      demo <- cort_sim2()
      p <- plot_cort_sim(demo)
      p
```

@@

@@

The goal of this simulation is to provide a flexible tool that can produce realistic datasets of physiological flexibility for a variety of different systems and scenarios. As such, there are many possible applications and here I briefly highlight a few possibilities. These are by no means exhaustive, and I hope the simulation will be a useful tool to guide empirical work for specific hypotheses and study systems. Within each scenario, I have illustrated how the simulation functions might be used to address the particular question of interest, but I have not fully explored all the possible permutations of parameters systematically, because these will depend to a large extent on the empirical details of the system being studied. A complete set of reproducible code to create all of the examples presented in this paper is available on GitHub (https://github.com/cct663/speed_vs_scope).

#### *Simulating empirically parameterized data*

@@

To demonstrate this utility, I have redrawn data from Koolhaas et al. [-@koolhaas2010]. As part of that study, a series of corticosterone measurements were collected during and after an acute stressor from 14 laboratory rats *Rattus norvegicus* using permanently implanted jugular vein canulae. I next simulated data using the functions described above starting with the input values calculated directly from the empirical data. The simulation creates a new dataset that has similar variation and patterns to the empirical data (Figure \@ref(fig:kool-demo)A) along with a population wide corticosterone response curve shape that closely matches the empirical data (Figure \@ref(fig:kool-demo)B). In this case, the plotted simulation data include the same number of animals sampled at the same time points as the empirical data, but these sampling points and total sample size can easily be changed as desired. The parameterized simulation can now be used to test the sensitivity of any number of experimental designs before additional data is collected.

#### *Accurately measuring a single glucocorticoid trait*

Single time point measures of glucocorticoids are often interpreted as representing meaningful variation between individuals. For example, variation in the level of glucocorticoids after 30 minutes of standardized restraint is typically interpreted as variation in the magnitude of the stress response [@taff2019]. However, this interpretation rests on assumptions that are rarely explicitly tested with empirical data. For example, the time chosen to take a stress-induced sample is often assumed to be either at the species peak or during a plateau period after the species peak. In some early studies, great care was taken to determine an average population level peak time [@wingfield1992], but many studies adopt the widely used 'standard' time of 30 minutes post capture without extensive validation [compiled in @vitousek2019]. While there is a general assumption that sampling later than the peak is acceptable (and perhaps preferable) because animals will be sampled during a relatively stable high plateau, there is little empirical data to evaluate this assertion or to determine how much under or overshooting the species peak timing might influence inferences. Furthermore, even when the average peak timing is well established, differences in the amount of between-individual variation in the time to reach the peak or in peak values are common across species and even in different life history stages within species [@wingfield1992]. The combinations of these patterns of variation could have major consequences on the accuracy of single point estimates taken at 30 minutes, but these questions cannot be addressed directly with empirical datasets where the true underlying values of each individual are unknown.

Here, I simulate a simple scenario exploring the consequences of variation in each of these parameters on the accuracy of estimating between individual differences in maximally expressed glucocorticoids during an acute response. For purposes of this illustration, I consider a single study design in which animals are sampled at 30 minutes. Using this design as a starting point, I systematically vary i) the timing of the population average peak (15, 30, or 45 minutes), ii) the amount of variation in maximum glucocorticoid levels reached, iii) and the amount of variation in the number of minutes taken to reach peak levels. All other variables in the simulation are constrained to be invariant between individuals in the population (e.g., all individuals have identical baseline glucocorticoids in this case), though I consider cases in which multiple aspects of the rapid response are correlated with each other in the next section. I included moderate within-individual variability and a small amount of assay error across all iterations. For each combination of parameters, I simulated 200 animals and estimated the R^2^ value from a regression of the observed estimates of glucocorticoid levels at 30 minutes to the true known values. This simulation is likely a best case scenario because it eliminates many sources of variation or noise that would be present in real data, but it illustrates the effect of variation in these three key parameters even when the exact same sampling design is employed.

@@

@@

#       f1 <- cort_sim2(data = cort_sim1(n = 50, max_mu = log(45), max_sd = 0.21, speed_mu = 30, speed_sd = 1), 

>>>>

---
title: "Simulating physiological flexibility in the acute glucocorticoid response to stressors reveals limitations of current empirical approaches"
author: 
 - Conor C. Taff
header-includes:
  - \usepackage[font={footnotesize}, labelfont={bf}]{caption}
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}\linenumbers
fontsize: 12pt
output: 
  bookdown::word_document2:
      number_sections: FALSE
      toc: FALSE
      extra_dependcies: ["flafter"]
  word_document:
  bookdown::pdf_document2:
      number_sections: FALSE
      toc: FALSE
      extra_dependencies: ["flafter"]
bibliography: references.bib
csl: animal-behaviour.csl
---

@@

Wild animals often experience unpredictable challenges that demand rapid and flexible responses. The glucocorticoid mediated stress response is one of the major systems that allows vertebrates to rapidly adjust their physiology and behavior. Given its role in responding to challenges, evolutionary physiologists have focused on the consequences of between-individual and, more recently, within-individual variation in the acute glucocorticoid response. However, empirical studies of physiological flexibility are severely limited by the logistical challenges of measuring the same animal multiple times. Data simulation is a powerful approach when empirical data are limited, but has not been adopted to date in studies of physiological flexibility. In this paper, I develop a simulation that can generate realistic acute glucocorticoid response data with user specified characteristics. Simulated animals can be sampled continuously through an acute response and across as many separate responses as desired, while varying key parameters. Using the simulation, I develop several scenarios that address key questions in physiological flexibility. These scenarios demonstrate the conditions under which a single glucocorticoid trait can be accurately assessed with typical experimental designs, the consequences of covariation between different components of the acute stress response, and the way that context specific differences in variability of acute responses can influence the power to detect relationships between the strength of the acute stress response and fitness. I also describe how to use the simulation tools to aid in the design and evaluation of empirical studies of physiological flexibility.

@@

Animals live in a dynamic environment in which they regularly encounter unpredictable challenges. Successfully navigating these challenges often requires the ability to rapidly adjust behavior and physiology to match current conditions. For vertebrates, the glucocorticoid mediated stress response plays a major role in coordinating these changes when stressors are encountered [@wingfield1998; @sapolsky2000] and similar rapid response systems mediate changes in other taxa [@taborsky2021]. Because of the central role that this response plays in coping with challenges, a great deal of research effort over the past 15 years has focused on understanding whether between-individual differences in the magnitude of this response predict coping ability and, ultimately, fitness [@breuner2008; @schoenle2020].

@@

Several recent papers have suggested that physiologists interested in endocrine flexibility should adopt a within-individual reaction norm approach [e.g., @hau2016; @taff2016]. This approach has been widely adopted in studies of behavioral flexibility where statistical methods and empirical progress have developed synergistically [e.g., @araya2015; @dingemanse2010; @westneat2015]. This field has also benefited from simulation studies to evaluate optimal study design [@van2012] and packages that can create artificial datasets with desired patterns of between, within, and residual variance to evaluate the consequences of different patterns of variation on the ability to detect effects [see SQuID package, @allegue2017]. While these approaches are powerful, they have proven difficult to apply directly to endocrine flexibility data for two reasons. First, simulation studies suggest that successfully modeling within-individual variation in flexible traits using an hierarchical modeling framework often requires a level of repeated sampling that is possible for many behaviors (especially when collected autonomously), but that is currently not possible for most studies of endocrine flexibility, because it would require sampling of many separate glucocorticoid responses per individual. Second, many behavioral papers focus on somewhat discrete measures (e.g., aggression score or activity level), whereas for acute glucocorticoid responses, the functional shape of the response itself may be the important trait. Fully describing the functional shape of a single acute glucocorticoid increase may require many samples in close succession, but for small vertebrates logistical and ethical constraints mean that it is rarely possible to take more than a few samples during the course of a single acute response.

The function valued trait (FVT) framework is an alternative approach that explicitly considers the functional shape of a biological response [@gomulkiewicz2018; @stinchcombe2012; @kingsolver2015]. While FVT approaches have been suggested for studies of endocrine flexibility, I am not aware of any papers that have applied this framework to empirical data on acute glucocorticoid responses, probably because sufficient data are not available. Conceptually, however, this approach is a better match to the acute glucocorticoid response, because the shape of a response curve is explicitly considered as the phenotypic trait of interest. In some cases, it may make sense to estimate particular parameters of the curve (e.g., maximum rate of increase and maximum value reached) and then treat those parameters as phenotypic values for downstream analysis, although statistical methods also exist to analyze the shape of the entire curve directly without the need to extract discrete parameters [@kingsolver2015]. This approach has been used to study a variety of phenotypes where values can be measured continuously or pooled across many individuals from the same group to accurately estimate the shape of a curve [see Table 1 in @stinchcombe2012]. Applying the technique to endocrine flexibility at the within-individual level faces the same empirical challenges described for within-individual reaction norms above, such as the need for repeated sampling of individuals and high temporal resolution of samples within individual physiological responses. Note that FVT and within-individual reaction norms approaches are not necessarily incompatible, but they have largely developed separately.

@@

INSERT HISTORY AND CITATIONS FROM COCKREM

For individual based studies, the most common approach to this problem is to standardize measurements as much as possible by measuring animals at the same time of the day during the same context, and by taking blood samples at standard times (often <3 and 30 minutes after capture) to characterize baseline and stress-induced glucocorticoids. This standardization allows for comparison between individuals, but in some cases it may also completely obscure the ability to detect variation in certain characteristics of the acute response curve. For example, if the speed (rate of initial increase or time required to reach maximum) and scope (maximum value) of the acute response vary independently, samples taken at only two time points cannot accurately capture variation in either parameter [@taff2022]. Indeed, several discussions in recent years about methods such as the '3 minute rule' and the relative merits of 'area under the curve' versus time point measures of glucocorticoids are fundamentally related to a recognition of the importance of understanding variation in the functional shape of stress responses and whether different components of that shape covary within individuals [e.g., @cockrem2002; @small2017]. While a great deal of empirical work has focused on characterizing the rate of initial increase [CITE], there has been relatively little work on understanding individual differences in the time required to reach maximal levels, and this attribute of speed is harder to estimate [@taff2022].

One of the characteristics of both the within-individual reaction norm and FVT literature is that empirical work has proceeded in very close coordination with simulation and statistical method development. In contrast, studies of endocrine flexibility often point to these methods, but don't address the ways that the particular logistical challenges of hormone measurement, such as the difficulty of repeatedly collecting samples within- or between-individuals, might necessitate different empirical approaches. I believe this is one reason that there are currently more conceptual papers arguing for a reaction norm approach to endocrine variation than there are empirical papers actually applying the approach [but see, @furtbauer2015; @taff2022; @lendvai2014; @houslay2022]. While many of the tools developed in these related fields are transferable, studies of physiological flexibility would benefit from a focus on analysis development and testing that explicitly incorporates the particular details and challenges of these questions. One way to accomplish these goals is to integrate empirical work with simulations, but to my knowledge no studies of physiological flexibility have developed simulations of the acute stress response that address the issues discussed above.

There is also a growing number of mathematical models of the  (Luttbeg, Taborsky)

@@

The goal of this simulation is to provide a flexible tool that can produce realistic datasets of physiological flexibility for a variety of different systems and scenarios. As such, there are many possible applications and here I briefly highlight a few possibilities. First, I demonstrate that the simulation can produce datasets that are qualitatively similar to empirical data on acute glucocorticoid responses. Next, I explore three specific scenarios with the simulation that highlight challenges to empirical progress in this field. First, I ask how well maximum glucocorticoid measures can be estimated with a single time point measure in populations that differ systematically in features of the response. Second, I explore how different patterns of covariation between maximum glucocorticoids and one measure of speed (the time to reach maximum) influence the ability to reliably measure either component of the response. Third, I ask how different amounts of between- and within-individual variation in maximum glucocorticoid levels impact the ability to detect known relationships between glucocorticoids and fitness. These questions are all important for understanding physiologically flexibility and are difficult to fully address empirically, but this is by no means an exhaustive list of the questions or possible parameter permutations that could be explored with simulation. Finally, I demonstrate how the simulation tools can be used to evaluate the performance of different experimental designs given a set of logistical limitations (e.g., number of samples that can be collected) and a proposed hypothesis) and discuss how this procedure could be used as a planning tool to increase the power and reproducibility of future empirical work.

@@

I developed a set of functions in R version 4.0.2 [@rcore] to generate acute physiological response curves. This simulation makes no assumptions about the  mechanistic process that results in the shape of a glucocorticoid response. Rather, parameters are sampled to generate curves that are similar in shape and degree of variation to empirically observed responses (Figure \@ref(fig:concept-fig)). This simulation is designed to create data sets with realistic structure that can be used to better design and plan studies of physiological flexibility, to evaluate power of current study designs, and to evaluate the sensitivity of sampling regimes to any number of modifications to the shape of glucocorticoid response curves (e.g., changing covariation patterns between different features of the response). I explore a small number of scenarios in the next section, but I expect that many other scenarios can be addressed with these tools. For illustration purposes, I refer to simulated glucocorticoid responses, but the simulation applies equally well to any physiological mediator of a rapid response. The code for the complete simulation along with full documentation of each function and argument can be accessed in a GitHub repository with a current version permanently archived on Zenodo (LINK). The package can be installed directly within the R environment using the following command.

@@

The simulation is constructed as two main functions with several minor functions for downstream analysis. Detailed descriptions of the arguments to each function are included with the package documentation. Briefly, function `cort_sim1` samples the parameters shown in Figure \@ref(fig:concept-fig) from an arbitrary number of animals. These parameters are sampled from a multivariate normal distribution with user specified mean, variance, and covariance for each parameter. By default, maximum glucocorticoid values are sampled from a normal distribution on the log scale and then exponentiated to determine absolute values. This results in a right skewed maximal glucocorticoid distribution on the absolute scale that is typical of many empirical datasets, but users can easily specify any parameters to be sampled from a normal or log normal distribution as required to match the characteristics of a particular study system. For the purposes of the simulation, I consider the sampled parameter values to be the 'true', unobserved, phenotype of the animal (setting aside the question of whether or not a 'true' physiological phenotype exists).

A second function, `cort_sim2`, starts with a population of animals generated from `cort_sim1` and samples observed acute glucocorticoid responses an arbitrary number of times for each animal. Two sources of variation in the observed relative to true parameter values can be specified. First, within-individual variation in expression is represented by specifying what amount of variation in the observation of each parameter is determined by the true value and what amount is determined by an additional randomly sampled response, based on the population parameters (this additional sampling maintains the user specified covariance structure of the population). After sampling the parameters, values are interpolated for each one minute time point and a localized regression is fit to create a smoothed curve that represents the observed glucocorticoid response. From this expressed response, individual data points are then collected at user specified times that would reflect an empirical study design (e.g., 1, 30, and 60 minutes). Additional noise can be added to these data points to represent measurement error (e.g., assay error). This simulated dataset can then be treated as the input for any desired analyses and statistical approaches, while maintaining the ability to compare results to the 'true' values used in the simulation.

The function also generates a simple simulated performance (e.g., fitness) measure, based on the underlying true parameter values sampled for each individual in the population (e.g., their baseline and maximum glucocorticoid value). The single fitness measure per animal is determined by allowing the user to specify the relative degree to which unmeasured traits plus each true parameter contribute to fitness outcomes. Data reflecting the true phenotypic values, the repeated expression of acute responses, and the observed time points can then be used in downstream analyses with any standard statistical approaches or software. For example, a user could perform an analysis to ask whether a known relationship between fitness and a particular true parameter is recovered in a study that includes only measures taken at particular time points. An additional convenience function summarizes the output of a simulation run in a multi-panel plot (Figure \@ref(fig:initial-demo)).

```{r initial-demo, results = FALSE, echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10.2, fig.height = 7.5, fig.cap = "Example of simulation output with default settings. Panel A shows the downsampled data set for this run with samples collected at 1, 15, and 30 minutes in this case. Panel B shows the full observed response curve for each animal. Panel C shows the rank order of glucocorticoid level at each time point for each animal. In each panel, the vertical dashed lines represent the three time points that might have been measured in a typical empirical study. Note that individuals in the top panels do not match perfectly because measurement error is added to the downsampled dataset in panel A."}
  # Make plot with default settings from package 
     set.seed(100)   # makes it reproducible by using same random seed
      demo <- cort_sim2()

          spoints <- unique(demo$simulated_dataset_long$time)
          p1 <- ggplot(data = demo$simulated_dataset, mapping = aes(x = time, 
              y = cort, by = animal_sample)) + geom_line(size = 0.5, 
              alpha = 0.7, color = "coral3") + guides(color = FALSE) + theme_classic() + 
              xlab("Time") + ylab("Corticosterone") + geom_vline(xintercept = spoints, 
              linetype = "dashed") +
              annotate(geom = "text", x = -Inf, y = Inf, hjust = -.5, vjust = 1.5, label = "A", size = 5)
          p2 <- ggplot(data = demo$timecourse_long, mapping = aes(x = time, 
              y = cort, color = animal, by = animal_sample)) + geom_line(size = 0.5, 
              color = "coral3", alpha = 0.7) + guides(color = FALSE) + 
              theme_classic() + xlab("Time") + ylab("Corticosterone") + 
              coord_cartesian(xlim = c(0, 60)) + geom_vline(xintercept = spoints, 
              linetype = "dashed") +
              annotate(geom = "text", x = -Inf, y = Inf, hjust = -.5, vjust = 1.5, label = "B", size = 5)
          p3 <- ggplot(data = demo$rank_timecourse, mapping = aes(x = time, 
              y = rank, color = animal_sample2)) + geom_line(size = 1.6, 
              alpha = 0.8) + guides(color = FALSE) + theme_classic() + 
              scale_color_viridis(discrete = TRUE) + xlim(0, 60) + 
              geom_vline(xintercept = spoints, linetype = "dashed") +
              annotate(geom = "text", x = -Inf, y = Inf, hjust = -.5, vjust = 1.5, label = "C", size = 5)

      grid.arrange(p1, p2, p3, layout_matrix = rbind(c(1, 2, 2), 
              c(3, 3, 3)))
```

@@

### METHODS FOR EACH SIMULATED SCENARIO

Each of the scenarios described here uses the basic simulation functions described above with input parameters adjusted to address the question of interest. A complete set of reproducible code to create all of the examples presented here is available on GitHub and permanenty archived at Zenodo (https://github.com/cct663/speed_vs_scope).

#### *Scenario 1: Simulating empirically parameterized data*

To demonstrate this utility, I attempted to create simulated datasets that recreate the characteristics of the empirical data presented in Koolhaas et al. [-@koolhaas2010]. As part of that study, a series of corticosterone measurements were collected during and after an acute stressor from 14 laboratory rats *Rattus norvegicus* using permanently implanted jugular vein canulae. I extracted data from Figure 6 in Koolhaas et al. [-@koolhaas2010] using WebPlotDigitizer (https://automeris.io/WebPlotDigitizer/) and then simulated data using the functions described above starting with the input values calculated directly from the empirical data. I then compared patterns of variation and population level response curves for the empirical and simulated data.

#### *Scenario 2: Accurately measuring a single glucocorticoid trait*

Single time point measures of glucocorticoids are often interpreted as representing meaningful variation between individuals. For example, variation in the level of glucocorticoids after 30 minutes of standardized restraint is typically interpreted as variation in the magnitude of the stress response [@taff2019]. However, this interpretation rests on assumptions that are rarely explicitly tested with empirical data. For example, the time chosen to take a stress-induced sample is often assumed to be either at the species peak or during a plateau period after the species peak. In some early studies, great care was taken to determine an average population level peak time [e.g., @wingfield1992], but many studies adopt the widely used 'standard' time of 30 minutes post capture without extensive validation [compiled in @vitousek2019]. While there is a general assumption that sampling later than the peak is acceptable (and perhaps preferable) because animals will be sampled during a relatively stable high plateau, there is little empirical data to evaluate this assertion or to determine how much under or overshooting the species peak timing might influence inferences. Furthermore, even when the average peak timing is well established, differences in the amount of between-individual variation in the time to reach the peak or in peak values are common across species and even in different life history stages within species [@wingfield1992]. Some studies focus instead on the rate of initial increase in glucocorticoids with a sample taken at 10 or 15 minutes after disturbance [e.g., @cockrem2005; @cockrem2006; @love2003], often well before peak levels are reached. The combination of different patterns of within- and between-individual variation with the exact time(s) chosen for sampling could have consequences for the accuracy of point estimates taken at any single time point, but these questions cannot be addressed directly with empirical datasets where the true underlying values of each individual are unknown.

I simulated a simple scenario to explore the consequences of variation in each of these parameters on the accuracy of estimating between individual differences in maximally expressed glucocorticoids during an acute response. For purposes of this illustration, I consider a single study design in which animals are sampled at 30 minutes. Using this design as a starting point, I systematically vary i) the timing of the population average peak (15, 30, or 45 minutes), ii) the amount of variation in maximum glucocorticoid levels reached (SD of 1 to 12 ng/$\mu$l), iii) and the amount of variation in the number of minutes taken to reach peak levels (SD of 1 to 20 minutes). All other variables in the simulation are constrained to be invariant between  individuals in the population (e.g., all individuals have identical baseline glucocorticoids in this case), though I consider cases in which multiple aspects of the rapid response are correlated with each other in the next scenario. I included moderate within-individual variability and a small amount of assay error across all iterations. For each combination of parameters, I simulated 200 animals and estimated the R^2^ value from a regression of the observed estimates of glucocorticoid levels at 30 minutes to the true known values. This simulation is likely a best case scenario for detection because it eliminates many sources of variation or noise that would be present in real data, but it illustrates the effect of variation in these three key parameters even when the exact same sampling design is employed.

#### *Scenario 3: Exploring covariance between response components*

#### *Scenario 4: Detecting links between fitness and responses*

Here, I imagine a simple scenario in which the 'true' maximum glucocorticoid level during an acute response explains 80% of the variation in fitness (clearly this is unrealistically high, but it is chosen for illustration only). I next construct a study in which researchers measure 50 individuals using a typical stress-induced (30 minute) sampling protocol. For simplicity, I set the other parameters in the simulation at their default values. Keeping the study design constant, I ask whether the glucocorticoid-fitness relationship can be recovered for two hypothetical populations that have low or high between-individual variation in maximum glucocorticoid levels. Using these populations, I asked asked how the ability to detect glucocorticoid-fitness relationships changed with different amounts of within-individual variation in acute response expression. For each combination of parameters, I simulated 50 populations and fit a simple linear regression model with observed glucocorticoid levels at 30 minutes as a predictor of fitness to ask whether the true glucocorticoid-fitness relationship was recovered.

#### *Scenario 5: Designing optimal sampling strategies*

One of the major benefits of simulating glucocorticoid response curves will be the ability to design optimal sampling strategies before data are collected. A simulation can be constrained to match any real world limitations (e.g., maximum number of samples possible per individual) and then explored to determine how to best allocate sampling resources. The specifics of this task will vary considerably with the study system and question being addressed, but here I illustrate one possible application. Consider an experiment in which the acute glucocorticoid response of a treatment group and control group are compared after some experimental manipulation. The details of the manipulation are unimportant here, but suppose that the prediction is that this manipulation should result in a difference in the speed of the corticosterone response between our two groups, such that the treatment group will reach it's maximum glucocorticoid value faster than the control group, but will not differ in the maximum value itself. I have implemented this difference by simulating two populations in which the treatment group has a steeper initial slope and also reaches the maximum value faster. Any number of possible hypotheses for a particular study system could be specified following a similar approach.

Using these simulated groups, I asked how well different study designs could detect the differences. Here we can impose any logistical constraints relevant to the study system. As an example, in this case we can only sample a maximum of 20 individuals per group, we can only sample each individual once post-treatment, and during that single sampling event we can take a blood sample at a maximum of three different time points, resulting in a total of 120 data points. Given these constraints, I compare several sampling designs. First, I explored 'standard' sampling strategies that are typical of empirical studies in this field that sample each individual animal at either i) 1 minute, 30 minutes, and 60 minutes, ii) 1 minute, 15 minutes, and 30 minutes, or iii) 1 minute, 15 minutes, and 60 minutes. Second, I explored two alternative sampling approaches that are not typically used in empirical work but might better capture the entire functional shape of response. These included, iv) a study in which three sampling times between 1 and 60 minutes are randomly chosen for every animal, and v) a study in which three sampling times are randomly chosen for each animal, but weighted more heavily around the range of times when maximum levels are expected to be reached for the population.

For illustration purposes I sampled directly from the 'true' response curves in this example so that there is no additional measurement error added. To evaluate these schemes I compare estimates of the acute response curve for each group to the 'true' known curves. Note that a more complete analysis of a sampling schemes performance should include many more iterations and full statistical comparisons, but the details here will be highly dependent on the study system and goals, so I provide this simple example to illustrate the approach rather than to make any more widely applicable conclusions.

@@

#### *Scenario 1: Simulating empirically parameterized data*

The simulation functions were able to produce a new synthetic dataset that has similar variation and patterns to the empirical data in Koolhaas et al. [-@koolhaas2010] (Figure \@ref(fig:kool-demo)A). The average population wide response curve shape also closely matched the empirical data (Figure \@ref(fig:kool-demo)B). In this case, the plotted simulation data include the same number of animals sampled at the same time points as the empirical data, but these sampling points and total sample size can easily be changed as desired. The parameterized simulation can now be used to test the sensitivity of any number of experimental designs before additional data is collected, such as different sample sizes or sampling time points.

@@

#### *Scenario 2: Accurately measuring a single glucocorticoid trait*

The amount of between-individual variation in maximum glucocorticoid values has a profound effect on the ability to detect true maximal levels with samples taken at 30 minutes \@ref(fig:max-sim). A single sample taken at 30 minutes was highly correlated with true maximal glucocorticoid levels when between-individual variation wass high, but with low between-individual variation in maximum a single sample was uninformative. I simulated a wide range of variation that may include unrealistically high and low values, but the result highlights the importance of considering the amount of variation expected in a study population. There is a weaker, but still substantial impact of variation in the time taken to reach maximum values on the accuracy of estimates in this simulation. Greater variation in the speed (time to reach maximum) of the response reduces the accuracy of estimates of maximal values. Finally, the timing of sampling relative to the average population peak timing also influences accuracy. Measuring after the average peak time results in the most accurate estimates across a range of parameter values, while measuring before the average peak time produces the least accurate measures, particularly when there is also high variation in the time to reach maximum values between individuals.

@@

# # Low within variation

@@

#       f1 <- simcoRt::cort_sim2(data = simcoRt::cort_sim1(n = 50, max_mu = log(45), max_sd = 0.21, speed_mu = 25, speed_sd = 1),
